---
title: "G2G Models with Time-Varying Covariates"
author: "Ka Lok Lee, Aaron Ramsey, and Noah Park"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-Varying Covariates with G2G Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{survival}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)
```

## Time-Varying Covariates in the G2G Model

The standard G2G model assumes that covariate values are fixed for each
subject. The time-varying extension replaces the static covariate effect
$C = \exp(\mathbf{x}'\boldsymbol{\beta})$ with a cumulative sum over
period-specific values:

$$C(t) = \sum_{s=1}^{t} \exp\!\left(\mathbf{x}(s)'\boldsymbol{\beta}\right)$$

The survivor function remains $S(t) = \bigl(1 + C(t)/\alpha\bigr)^{-r}$, but
now evolves with the covariate path of each subject. This accommodates
marketing interventions, clinical treatment changes, or any predictor that
varies week-to-week or period-to-period.

### Required Data Format

Data must be in **person-period (long) format**:

- One row per subject per time period
- A subject identifier column linking rows to the same individual
- The event indicator equals 1 in the period the event occurred, 0 in all
  other periods (including the final period for censored subjects)

## Setup

```{r load-packages}
library(G2Gcov)
library(survival)
```

## Example Data: Customer Behavior (`kb_data`)

The `kb_data` dataset included in the package tracks weekly covariate values
for a panel of customers. Each row is one customer-week.

```{r load-data}
data(kb_data, package = "G2Gcov")
head(kb_data)
```

| Variable    | Description |
|-------------|-------------|
| `id`        | Customer identifier |
| `week`      | Week number |
| `num_weeks` | Total weeks the customer was observed |
| `censor`    | 1 = still active (censored), 0 = churned (event) |
| `coupon`    | Coupon stock covariate (time-varying) |
| `anyp`      | Trade-promotions covariate (time-varying) |

## Data Preparation

`censor = 0` indicates the event (churn) occurred in that week. The model
requires a `status` column equal to `1 - censor`:

```{r prepare-data}
kb_data$status <- 1L - kb_data$censor

# Confirm format: one row per (id, week), unique
stopifnot(!anyDuplicated(kb_data[, c("id", "week")]))

# Summary at the subject level
n_subjects <- length(unique(kb_data$id))
n_events   <- sum(
  tapply(kb_data$status, kb_data$id, max)  # 1 if subject ever had status=1
)

cat("Subjects:        ", n_subjects, "\n")
cat("Churned:         ", n_events, "\n")
cat("Censored:        ", n_subjects - n_events, "\n")
cat("Total obs. rows: ", nrow(kb_data), "\n")
```

## Fitting the Model

`G2G_varying_MLE()` takes a `Surv()` formula, the long-format data frame, and
the name of the subject ID column.

```{r fit-full}
fit_full <- G2G_varying_MLE(
  fo      = Surv(week, status) ~ coupon + anyp,
  data    = kb_data,
  subject = "id"
)

cat("Convergence code:", fit_full$convergence,
    ifelse(fit_full$convergence == 0, "(converged)", "(check starting values)"), "\n")
cat("Negative log-likelihood:", round(fit_full$value, 2), "\n")
```

## Parameter Estimates

```{r results-table}
results <- data.frame(
  Parameter = names(fit_full$par),
  Estimate  = fit_full$par,
  SE        = fit_full$par_stderr,
  CI_Lower  = fit_full$par_lower,
  CI_Upper  = fit_full$par_upper,
  row.names = NULL
)

knitr::kable(
  results,
  digits    = 3,
  col.names = c("Parameter", "Estimate", "Std. Error",
                "Lower 95% CI", "Upper 95% CI"),
  caption   = "G2G parameter estimates -- kb_data (full model)"
)
```

**Interpreting the estimates:**

- **r** and **alpha** define the Gamma distribution over baseline churn
  propensity. Mean baseline propensity = $r / \alpha$.
- **coupon** and **anyp** are log-hazard coefficients. A positive coefficient
  means higher covariate values increase the churn hazard in that period.

### Hazard Ratios

```{r hazard-ratios}
coef_idx <- seq_along(fit_full$par)[-(1:2)]
hr_df <- data.frame(
  Covariate   = names(fit_full$par)[coef_idx],
  Coefficient = fit_full$par[coef_idx],
  HR          = exp(fit_full$par[coef_idx]),
  Direction   = ifelse(fit_full$par[coef_idx] > 0,
                       "Increases churn", "Decreases churn"),
  row.names   = NULL
)
knitr::kable(
  hr_df,
  digits    = 3,
  col.names = c("Covariate", "Coefficient", "Hazard Ratio", "Direction"),
  caption   = "Covariate hazard ratios"
)
```

A hazard ratio greater than 1 means a one-unit increase in that covariate
multiplies the period churn hazard by that factor.

## Model Comparison

### Fit Nested Models

```{r fit-reduced}
fit_coupon <- G2G_varying_MLE(
  fo = Surv(week, status) ~ coupon, data = kb_data, subject = "id"
)
fit_anyp <- G2G_varying_MLE(
  fo = Surv(week, status) ~ anyp,   data = kb_data, subject = "id"
)
```

### Log-Likelihood Comparison

```{r compare-models}
mc <- data.frame(
  Model        = c("Full (coupon + anyp)", "Coupon only", "Anyp only"),
  LogLik       = c(-fit_full$value, -fit_coupon$value, -fit_anyp$value),
  N_Parameters = c(length(fit_full$par), length(fit_coupon$par),
                   length(fit_anyp$par)),
  row.names    = NULL
)
knitr::kable(
  mc,
  digits    = 2,
  col.names = c("Model", "Log-Likelihood", "Parameters"),
  caption   = "Model comparison by log-likelihood"
)
```

### Likelihood Ratio Tests

```{r lrt}
# LRT: full vs. model without one covariate
lrt_coupon <- 2 * ((-fit_anyp$value)   - (-fit_full$value))
lrt_anyp   <- 2 * ((-fit_coupon$value) - (-fit_full$value))

lrt_df <- data.frame(
  Covariate = c("coupon", "anyp"),
  LRT       = c(lrt_coupon, lrt_anyp),
  df        = c(1L, 1L),
  p_value   = c(pchisq(lrt_coupon, 1, lower.tail = FALSE),
                pchisq(lrt_anyp,   1, lower.tail = FALSE)),
  row.names = NULL
)
knitr::kable(
  lrt_df,
  digits    = c(0, 3, 0, 4),
  col.names = c("Dropped covariate", "LRT statistic", "df", "p-value"),
  caption   = "Likelihood ratio tests (full vs. reduced)"
)
```

A small p-value indicates that removing that covariate significantly worsens
model fit.

## Visualizations

### Baseline Propensity Distribution

```{r plot-p0, fig.cap = "Fitted Gamma distribution over baseline churn propensity"}
plot_p0(fit_full)
```

### In-Sample and Hold-Out Survival

`plot_insample_holdout_paper()` splits the data at a chosen week, fits the
model on the training period, and then plots observed (Kaplan-Meier) versus
G2G-predicted survival and retention rates across both periods. The dashed
vertical line marks the train/test boundary.

```{r plot-paper, fig.cap = "Actual vs. G2G predicted survival (left) and retention (right)"}
out <- plot_insample_holdout_paper(
  fo       = Surv(week, status) ~ coupon + anyp,
  data     = kb_data,
  subject  = "id",
  split_at = 12L
)

print(out$p_survival)
print(out$p_retention)
```

Good model fit is indicated by the dashed G2G line tracking closely with
the solid observed line, both within the training window and in the hold-out
period beyond week 12.

## References

Fader, P.S. and Hardie, B.G.S. (1997). How to project customer retention.
*Journal of Interactive Marketing*, 11(1), 61--73.

Fader, P.S. and Hardie, B.G.S. (2007). How to project customer retention
revisited: The role of duration dependence.
*Journal of Interactive Marketing*, 21(1), 76--90.
<https://doi.org/10.1002/dir.20074>

Grassia, A. (1977). On a family of distributions with argument between 0 and
1 obtained by transformation of the Gamma distribution and derived compound
distributions. *Australian Journal of Statistics*, 19(2), 108--114.
<https://doi.org/10.1111/j.1467-842X.1977.tb01277.x>

Schmittlein, D.C., Morrison, D.G. and Colombo, R. (1987). Counting your
customers: Who are they and what will they do next?
*Management Science*, 33(1), 1--24.
<https://doi.org/10.1287/mnsc.33.1.1>

## Session Information

```{r session-info}
sessionInfo()
```
