---
title: "G2G Models with Time-Varying Covariates"
author: "Ka Lok Lee, Aaron Ramsey, and Noah Park"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-Varying Covariates with G2G Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{survival, eha}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)
```

## Time-Varying Covariates in the G2G Model

The standard G2G model assumes that covariate values are fixed for each
subject. The time-varying extension replaces the static covariate effect
$C = \exp(\mathbf{x}'\boldsymbol{\beta})$ with a cumulative sum over
period-specific values:

$$C(t) = \sum_{s=1}^{t} \exp\!\left(\mathbf{x}(s)'\boldsymbol{\beta}\right)$$

The survivor function remains $S(t) = \bigl(1 + C(t)/\alpha\bigr)^{-r}$, but
now evolves with the covariate path of each subject. This accommodates
marketing interventions, clinical treatment changes, or any predictor that
varies week-to-week or period-to-period.

### Required Data Format

Data must be in **person-period (long) format**:

- One row per subject per time period
- A subject identifier column linking rows to the same individual
- The event indicator equals 1 in the period the event occurred, 0 in all
  other periods (including the final period for censored subjects)

---

## Example 1: Customer Churn (`kb_data`)

The `kb_data` dataset included in the package tracks weekly covariate values
for a panel of customers. Each row is one customer-week. This example shows
the typical marketing application: do coupon activity and trade promotions
affect the weekly probability of churn?

```{r load-packages-1}
library(G2Gcov)
library(survival)
```

```{r load-kb}
kb_data <- G2Gcov::kb_data
head(kb_data)
```

| Variable    | Description |
|-------------|-------------|
| `id`        | Customer identifier |
| `week`      | Week number |
| `num_weeks` | Total weeks observed |
| `censor`    | 1 = still active (censored), 0 = churned (event) |
| `coupon`    | Coupon stock covariate (time-varying) |
| `anyp`      | Trade-promotions covariate (time-varying) |

### Data Preparation

`censor = 0` indicates the event occurred. The model requires a `status`
column equal to `1 - censor`:

```{r prepare-kb}
kb_data$status <- 1L - kb_data$censor

n_subjects <- length(unique(kb_data$id))
n_events   <- sum(tapply(kb_data$status, kb_data$id, max))

cat("Subjects:        ", n_subjects, "\n")
cat("Churned:         ", n_events, "\n")
cat("Censored:        ", n_subjects - n_events, "\n")
cat("Total obs. rows: ", nrow(kb_data), "\n")
```

### Fitting the Model

```{r fit-kb}
fit_full <- G2G_varying_MLE(
  fo      = Surv(week, status) ~ coupon + anyp,
  data    = kb_data,
  subject = "id"
)

cat("Convergence:", fit_full$convergence,
    ifelse(fit_full$convergence == 0, "(converged)", "(check starting values)"), "\n")
cat("Negative log-likelihood:", round(fit_full$value, 2), "\n")
```

### Parameter Estimates

```{r results-kb}
results <- data.frame(
  Parameter = names(fit_full$par),
  Estimate  = fit_full$par,
  SE        = fit_full$par_stderr,
  CI_Lower  = fit_full$par_lower,
  CI_Upper  = fit_full$par_upper,
  row.names = NULL
)
knitr::kable(
  results,
  digits    = 3,
  col.names = c("Parameter", "Estimate", "Std. Error",
                "Lower 95% CI", "Upper 95% CI"),
  caption   = "G2G parameter estimates -- kb_data (full model)"
)
```

**Interpreting the estimates:**

- **r** and **alpha** define the Gamma distribution over baseline churn
  propensity. Mean baseline propensity = $r / \alpha$.
- **coupon** and **anyp** are log-hazard coefficients. A positive coefficient
  means higher covariate values increase the churn hazard in that period.

### Hazard Ratios

```{r hr-kb}
coef_idx <- seq_along(fit_full$par)[-(1:2)]
hr_df <- data.frame(
  Covariate   = names(fit_full$par)[coef_idx],
  Coefficient = fit_full$par[coef_idx],
  HR          = exp(fit_full$par[coef_idx]),
  Direction   = ifelse(fit_full$par[coef_idx] > 0,
                       "Increases churn", "Decreases churn"),
  row.names   = NULL
)
knitr::kable(
  hr_df,
  digits    = 3,
  col.names = c("Covariate", "Coefficient", "Hazard Ratio", "Direction"),
  caption   = "Covariate hazard ratios"
)
```

### Model Comparison and Likelihood Ratio Tests

```{r lrt-kb}
fit_coupon <- G2G_varying_MLE(
  fo = Surv(week, status) ~ coupon, data = kb_data, subject = "id"
)
fit_anyp <- G2G_varying_MLE(
  fo = Surv(week, status) ~ anyp,   data = kb_data, subject = "id"
)

lrt_coupon <- 2 * ((-fit_anyp$value)   - (-fit_full$value))
lrt_anyp   <- 2 * ((-fit_coupon$value) - (-fit_full$value))

lrt_df <- data.frame(
  Covariate = c("coupon", "anyp"),
  LRT       = c(lrt_coupon, lrt_anyp),
  df        = c(1L, 1L),
  p_value   = c(pchisq(lrt_coupon, 1, lower.tail = FALSE),
                pchisq(lrt_anyp,   1, lower.tail = FALSE)),
  row.names = NULL
)
knitr::kable(
  lrt_df,
  digits    = c(0, 3, 0, 4),
  col.names = c("Dropped covariate", "LRT statistic", "df", "p-value"),
  caption   = "Likelihood ratio tests (full vs. reduced)"
)
```

### Visualizations

```{r plot-p0-kb, fig.cap = "Gamma distribution over baseline churn propensity"}
plot_p0(fit_full)
```

```{r plot-paper-kb, fig.cap = "Actual vs. G2G predicted survival and retention (kb_data)"}
out_kb <- plot_insample_holdout_paper(
  fo       = Surv(week, status) ~ coupon + anyp,
  data     = kb_data,
  subject  = "id",
  split_at = 12L
)
print(out_kb$p_survival)
print(out_kb$p_retention)
```

The dashed vertical line marks the train/test boundary at week 12. G2G
predictions beyond this line are hold-out forecasts based solely on parameters
estimated from the training window.

---

## Example 2: Historical Mortality (Scania)

This example demonstrates that `G2G_varying_MLE()` generalizes beyond
marketing data to historical demographic data. The `scania` dataset from the
`eha` package (Broström, 2012) records individual mortality histories in
19th-century Sweden. It combines two time-invariant covariates (`sex`,
`immigrant`) with one genuinely time-varying covariate: the annual log rye
price (`foodprices`), a proxy for nutritional stress that changes each
calendar year. This makes it a natural application for the time-varying G2G
model.

```{r load-eha}
scania <- eha::scania
logrye <- eha::logrye
```

### Data Description

The `scania` dataset is in individual-level (short) format — one row per
person. The `logrye` dataset contains one row per calendar year with the
log rye price for that year.

| Variable     | Source    | Type           | Description |
|--------------|-----------|----------------|-------------|
| `id`         | `scania`  | Time-invariant | Individual identifier |
| `sex`        | `scania`  | Time-invariant | Sex (female / male) |
| `immigrant`  | `scania`  | Time-invariant | Immigrant status (yes / no) |
| `event`      | `scania`  | Outcome        | 1 = died, 0 = censored |
| `foodprices` | `logrye`  | Time-varying   | Log rye price for that calendar year |

### Data Preparation

The `scania` data is in individual-level (short) format and must be converted
to person-period (long) format before fitting. We center observation time at
age 50 so that period 1 represents the first year of risk at or after age 50.
Each individual's `spell` is their total years of observation.

```{r prepare-scania-individual}
sc <- scania
sc$exit      <- ceiling(sc$exit)
sc$birthdate <- floor(sc$birthdate)
sc$spell     <- sc$exit - sc$enter
sc$enter     <- sc$enter - 50L
sc$exit      <- sc$exit  - 50L
sc           <- sc[sc$spell >= 1L, ]

cat("Individuals:", nrow(sc), "  Events:", sum(sc$event), "\n")
```

The full dataset has over 12,000 individuals. We draw a reproducible sample
of 500 subjects so this vignette builds in reasonable time. In practice,
fit on the full dataset.

```{r sample-scania}
set.seed(123)
ids_sample <- sample(unique(sc$id), 500L)
sc_sample  <- sc[sc$id %in% ids_sample, ]
```

We expand to person-period format using a base R helper, then join the annual
food price. Each subject gets one row per year of observation. The event
indicator is 1 only in the final row for subjects who died; all other rows
are 0.

```{r prepare-scania-panel}
# Expand individual-level data to one row per subject per period
to_person_period <- function(data, time_col, event_col) {
  rows <- lapply(seq_len(nrow(data)), function(i) {
    spell <- data[[time_col]][i]
    if (is.na(spell) || spell < 1L) return(NULL)
    df         <- data[rep(i, spell), , drop = FALSE]
    df$timeInt <- seq_len(spell)
    df$y       <- c(rep(0L, spell - 1L), as.integer(data[[event_col]][i]))
    row.names(df) <- NULL
    df
  })
  do.call(rbind, rows)
}

sc_pp        <- to_person_period(sc_sample, time_col = "spell",
                                 event_col = "event")
sc_pp$event  <- sc_pp$y
sc_pp$exit   <- sc_pp$timeInt
sc_pp$enter  <- sc_pp$exit - 1L
sc_pp$age    <- sc_pp$enter + 50L

# Calendar year = birth year + age; join annual food prices
sc_pp$year   <- sc_pp$age + floor(sc_pp$birthdate)
sc_pp        <- merge(sc_pp, logrye, by = "year", all.x = TRUE)
sc_pp        <- sc_pp[, c("id", "enter", "exit", "event",
                           "sex", "immigrant", "age", "year", "foodprices")]

cat("Person-period rows:", nrow(sc_pp), "\n")
head(sc_pp)
```

### Fitting the Model

We fit a G2G model with two time-invariant covariates (`sex`, `immigrant`)
and the time-varying `foodprices`. The model estimates how each factor
shifts the annual mortality hazard, while the Gamma frailty (r, alpha)
captures unobserved heterogeneity in baseline mortality propensity across
individuals.

```{r fit-scania}
fit_sc <- G2G_varying_MLE(
  fo      = Surv(exit, event) ~ sex + immigrant + foodprices,
  data    = sc_pp,
  subject = "id"
)

cat("Convergence:", fit_sc$convergence,
    ifelse(fit_sc$convergence == 0, "(converged)", "(check starting values)"), "\n")
cat("Negative log-likelihood:", round(fit_sc$value, 2), "\n")
```

### Parameter Estimates

```{r results-scania}
results_sc <- data.frame(
  Parameter = names(fit_sc$par),
  Estimate  = fit_sc$par,
  SE        = fit_sc$par_stderr,
  CI_Lower  = fit_sc$par_lower,
  CI_Upper  = fit_sc$par_upper,
  row.names = NULL
)
knitr::kable(
  results_sc,
  digits    = 3,
  col.names = c("Parameter", "Estimate", "Std. Error",
                "Lower 95% CI", "Upper 95% CI"),
  caption   = "G2G parameter estimates -- Scania sample"
)
```

### Hazard Ratios

Exponentiate covariate coefficients to obtain hazard ratios — the
multiplicative change in period mortality hazard for a one-unit increase in
each predictor.

```{r hr-scania}
coef_idx_sc <- seq_along(fit_sc$par)[-(1:2)]
hr_sc <- data.frame(
  Covariate   = names(fit_sc$par)[coef_idx_sc],
  Coefficient = fit_sc$par[coef_idx_sc],
  HR          = exp(fit_sc$par[coef_idx_sc]),
  row.names   = NULL
)
knitr::kable(
  hr_sc,
  digits    = 3,
  col.names = c("Covariate", "Coefficient", "Hazard Ratio"),
  caption   = "Covariate hazard ratios -- Scania"
)
```

A hazard ratio above 1 increases mortality risk; below 1 decreases it. The
`foodprices` coefficient captures whether higher annual food prices — a proxy
for nutritional scarcity — were associated with elevated mortality in that
year, consistent with historical evidence on famine and mortality in
19th-century Sweden.

### Visualizations

The baseline propensity distribution shows the Gamma-distributed heterogeneity
in underlying mortality risk across individuals, after accounting for sex,
immigrant status, and food prices.

```{r plot-p0-scania, fig.cap = "Gamma distribution over baseline mortality propensity (Scania)"}
plot_p0(fit_sc)
```

The in-sample survivor curve plots the mean predicted S(t) — the probability
of surviving to each period — averaged across all subjects in the sample.

```{r plot-survival-scania, fig.cap = "In-sample G2G mean survivor curve (Scania)"}
plot_survival_insample_g2g(
  fo         = Surv(exit, event) ~ sex + immigrant + foodprices,
  train      = sc_pp,
  subject    = "id",
  model_type = "varying"
)
```

---

## References

Broström, G. (2012). *Event History Analysis with R*. Chapman & Hall/CRC.

Fader, P.S. and Hardie, B.G.S. (1997). How to project customer retention.
*Journal of Interactive Marketing*, 11(1), 61--73.

Fader, P.S. and Hardie, B.G.S. (2007). How to project customer retention
revisited: The role of duration dependence.
*Journal of Interactive Marketing*, 21(1), 76--90.
<https://doi.org/10.1002/dir.20074>

Grassia, A. (1977). On a family of distributions with argument between 0 and
1 obtained by transformation of the Gamma distribution and derived compound
distributions. *Australian Journal of Statistics*, 19(2), 108--114.
<https://doi.org/10.1111/j.1467-842X.1977.tb01277.x>

Schmittlein, D.C., Morrison, D.G. and Colombo, R. (1987). Counting your
customers: Who are they and what will they do next?
*Management Science*, 33(1), 1--24.
<https://doi.org/10.1287/mnsc.33.1.1>

## Session Information

```{r session-info}
sessionInfo()
```
