---
title: "G2G Models with Time-Invariant Covariates"
author: "Ka Lok Lee, Aaron Ramsey, and Noah Park"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{G2G Models with Time-Invariant Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{survival}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  warning  = FALSE,
  message  = FALSE
)
```

## The G2G Model

The **Grassia(II)-Geometric (G2G)** model is a discrete-time survival model that
accounts for unobserved heterogeneity across subjects (Fader and Hardie, 1997).
Each subject has a latent event propensity drawn from a Gamma distribution:

$$p_0 \sim \text{Gamma}(r,\, \alpha)$$

Covariates shift this baseline propensity through a log-linear link. For a
subject with covariate vector **x**, the cumulative covariate effect at period
*t* is $C(t) = t \cdot \exp(\mathbf{x}'\boldsymbol{\beta})$, and the survivor
function is:

$$S(t) = \left(1 + \frac{C(t)}{\alpha}\right)^{-r}$$

The parameters **r** (shape) and **alpha** (rate) characterize the population
distribution of baseline propensity; **beta** is a vector of covariate
log-hazard coefficients.

## Setup

```{r load-packages}
library(G2Gcov)
library(survival)
```

## Example Data: Veteran Lung Cancer Trial

We use the `veteran` dataset from the `survival` package (Kalbfleisch and
Prentice, 1980), a randomized trial of two treatments for lung cancer with
137 patients.

```{r load-data}
# veteran is lazy-loaded in the survival package; accessible directly
vet <- survival::veteran

# Add a unique subject ID (required by the visualisation functions)
vet$id <- seq_len(nrow(vet))

head(vet[, c("id", "time", "status", "age", "karno")])
```

| Variable | Description |
|----------|-------------|
| `time`   | Survival time (days) |
| `status` | 1 = death, 0 = censored |
| `age`    | Age in years |
| `karno`  | Karnofsky performance score (0–100; higher is better) |

```{r data-summary}
cat("Subjects:", nrow(vet), "\n")
cat("Events:  ", sum(vet$status), "\n")
cat("Censored:", sum(1 - vet$status), "\n")
```

## Fitting the Model

`G2G_static_MLE()` accepts a standard `Surv()` formula. Internally it
maximizes the G2G log-likelihood via BFGS and returns parameter estimates,
standard errors, and 95% confidence intervals.

```{r fit-model}
fit <- G2G_static_MLE(
  Surv(time, status) ~ age + karno,
  data = vet
)

cat("Convergence code:", fit$convergence,
    ifelse(fit$convergence == 0, "(converged)", "(check starting values)"), "\n")
cat("Negative log-likelihood:", round(fit$value, 2), "\n")
```

## Parameter Estimates

```{r results-table}
results <- data.frame(
  Parameter = names(fit$par),
  Estimate  = fit$par,
  SE        = fit$par_stderr,
  CI_Lower  = fit$par_lower,
  CI_Upper  = fit$par_upper,
  row.names = NULL
)

knitr::kable(
  results,
  digits    = 3,
  col.names = c("Parameter", "Estimate", "Std. Error",
                "Lower 95% CI", "Upper 95% CI"),
  caption   = "G2G parameter estimates -- veteran data"
)
```

**Interpreting the estimates:**

- **r (shape)** and **alpha (rate)** define the Gamma distribution over
  baseline event propensity. The population mean propensity is
  $r / \alpha = `r round(fit$par[1] / fit$par[2], 3)`$.
- Covariate coefficients are on the log-hazard scale. Positive values
  increase the hazard (shorter survival); negative values decrease it.

**Hazard ratios** are obtained by exponentiating covariate coefficients:

```{r hazard-ratios}
coef_idx <- seq_along(fit$par)[-(1:2)]
hr_df <- data.frame(
  Covariate   = names(fit$par)[coef_idx],
  Coefficient = fit$par[coef_idx],
  HR          = exp(fit$par[coef_idx]),
  row.names   = NULL
)
knitr::kable(
  hr_df,
  digits    = 3,
  col.names = c("Covariate", "Coefficient", "Hazard Ratio"),
  caption   = "Covariate hazard ratios"
)
```

A hazard ratio below 1 indicates a protective effect. Higher Karnofsky score
(better functional status) is associated with reduced hazard, consistent with
clinical expectation.

## Baseline Propensity Distribution

`plot_p0()` plots the fitted Gamma density over baseline event propensity.
The shaded region covers the central 90% of the distribution (default:
central 90% band); the solid vertical line marks the mean.

```{r plot-p0, fig.cap = "Fitted Gamma distribution over baseline event propensity"}
plot_p0(fit)
```

A wide distribution indicates substantial unobserved heterogeneity; a narrow
distribution suggests a relatively homogeneous population.


## Survivor Curve

`plot_survival_insample_g2g()` plots the mean predicted survivor function
S(t) over the training data.

```{r plot-survival, fig.cap = "In-sample G2G mean survivor curve"}
plot_survival_insample_g2g(
  fo         = Surv(time, status) ~ age + karno,
  train      = vet,
  subject    = "id",
  model_type = "static"
)
```

## Model Diagnostics

```{r diagnostics}
# AIC: 2k - 2 * log-likelihood
k   <- length(fit$par)
aic <- 2 * k + 2 * fit$value
cat("Parameters:", k, "\n")
cat("AIC:       ", round(aic, 2), "\n")

# Standard errors (NA when Hessian is singular)
if (anyNA(fit$par_stderr)) {
  cat("Note: some standard errors are NA (near-singular Hessian).\n")
} else {
  cat("All standard errors are finite.\n")
}
```

## References

Fader, P.S. and Hardie, B.G.S. (1997). How to project customer retention.
*Journal of Interactive Marketing*, 11(1), 61--73.

Grassia, A. (1977). On a family of distributions with argument between 0 and
1 obtained by transformation of the Gamma distribution and derived compound
distributions. *Australian Journal of Statistics*, 19(2), 108--114.
<https://doi.org/10.1111/j.1467-842X.1977.tb01277.x>

Kalbfleisch, J.D. and Prentice, R.L. (1980).
*The Statistical Analysis of Failure Time Data*.
Wiley, New York.

## Session Information

```{r session-info}
sessionInfo()
```
